---
title: "Data Drive Management II - Lab 5"
author: "Kyle Crichton"
date: "February 17, 2017"
output: github_document
---

#Setup

Load Packages

```{r, warning=F, message=F }
library(RCurl)
library(sp)
library(maptools)
library(censusapi)
library(ggmap)
```

#Housing Data

##Load Home Price Dataset

Store URL for data, read in as a csv and store data as  data frrame.

```{r, warning=F, message=F }
my.url <- "https://docs.google.com/spreadsheets/d/1W0vM5sCRhZjkQh6A0WGV8j1rhQAjecCrQW7PguHft-E/pub?gid=1989082857&single=true&output=csv"

dat.raw <- getURL( my.url, ssl.verifypeer=FALSE )

dat <- read.csv( textConnection(dat.raw), stringsAsFactors=FALSE )
```

Rename data frame columns

```{r, warning=F, message=F }
names(dat) <- c("timestamp","price","X1","X2","sqft","your.name","lot.size","beds","bath","garage","year","elementary","middle","high","walk","tax","highway","restaurant","starbucks","park","mall","address","zip","tract" )
```

Sample of the dataset is below:

```{r, warning=F, message=F, echo=F }
head(dat)
```

##Format Dataset for Analysis

Remove commas from strings and convert to numerics.

```{r, warning=F, message=F}
dat$price <- as.numeric( gsub( ",","", dat$price ) )
dat$tax <- as.numeric( gsub( ",","", dat$tax ) )
dat$lot.size <- as.numeric( gsub( ",","", dat$lot.size ) )
dat$sqft <- as.numeric( gsub( ",","", dat$sqft ) )
dat$lot.size[ is.na( dat$lot.size ) ] <- mean( dat$lot.size, na.rm=T )
```

#Geocode House Data

##Data Cleansing on Addresses

Some of the addresses stored in the address field are just street addresses, but some contain the city, state and zipcode in addition to the street address.  We want complete street addresses when geocoding the data but in order to standardize the formats we will first remove the extra information from the fields before adding the city, state and zipcode back in.

Example of the unstandardized data:

```{r, warning=F, message=F, echo=F }
dat[122:126, "address"]
```

Remove the strings ", Syracuse NY", ", Syracuse, NY", "Syracuse NY" and "Syracuse, NY"

```{r, warning=F, message=F }
dat$address <- sub(", Syracuse NY", "", dat$address)
dat$address <- sub(", Syracuse, NY", "", dat$address)
dat$address <- sub("Syracuse NY", "", dat$address)
dat$address <- sub("Syracuse, NY", "", dat$address)
```

Remove regular expression patterns matching 5 digits at the end of a string (zipcodes).

```{r, warning=F, message=F }
dat$address <- sub("[0-9]{5}$", "", dat$address)
```

Trim leftover trailing whitespace

```{r, warning=F, message=F }
dat$address <- sub("\\s+$", "", dat$address)
```

Fix bad zipcode (change 1323 to 13203)

```{r, warning=F, message=F }
dat$zip <- sub("1323", "13203", dat$zip)
```

Fixed Data:

```{r, warning=F, message=F, echo=F }
dat[122:126, "address"]
```

##Concatenate Address Fields

Pulls together the standardized address field, appends the city (Syracuse) and state (NY), and then concats the zipcode field.

```{r, warning=F, message=F }
dat$full.addr <- paste(dat$address, ",", sep="")
dat$full.addr <- paste(dat$full.addr, "Syracuse, NY", dat$zip, sep=" ")
```

Sample of full address data:

```{r, warning=F, message=F, echo=F }
dat[122:126, "full.addr"]
```

##Geocode Addresses

```{r, warning=F, message=F }
options( warn=-1 ) 

suppressMessages( dat$coords <- geocode(dat$full.addr, messaging=F ))
```

#Shapefiles

Read in the census tract shapefiles for Syracuse.

```{r, warning=F, message=F }
syr <- readShapePoly(fn="C:/Users/krcrichton/Documents/R/Support_Files/tl_2015_36_tract/tl_2015_36_tract",proj4string=CRS("+proj=longlat +datum=WGS84"))
```

#Spatial Join - Census Tract

Convert spatial coordinates to match shapefiles projection and match house points to census tract polygons.

```{r, warning=F, message=F }
houses <- SpatialPoints(dat$coords, proj4string=CRS("+proj=longlat +datum=WGS84"))
tracts.houses <- over(houses, syr)
dat.fips <- tracts.houses[,"GEOID"]
```

Map house locations

```{r, warning=F, message=F, fig.width = 10 }
plot(syr, border="white", col="grey90", main="Location of Homes in Dataset", xlim=c(-76.2, -76.1), ylim=c(43.0, 43.1))
points(dat$coords, pch=20, cex = 0.7, col="darkslategrey")
```

Store FIPS ID in house dataset.

```{r, warning=F, message=F }
dat$fips <- tracts.houses$GEOID
```

Sample of FIPS IDs:

```{r, warning=F, message=F, echo=F }
head(dat$fips)
```

#Census Data

##Retrieve Census Data

Store censusapi key to be used in the api call.

```{r, warning=F, message=F }
cKey <- "c4a7ded6e9442cdb7584465e73b4aff65aeaff8e"
```

Get census data and limit to Syracuse area.

```{r, warning=F, message=F }
census.dat <- getCensus(name="acs5", vintage=2015, key=cKey, 
                  vars=c("B08136_001E", "B23025_005E", "B19301_001E"), 
                  region="tract:*", regionin="state:36+county:067")
```

Rename data frame columns to be logical

B08136_001E is Travel Time to Work
B23025_005E is Unemployed
B19301_001E is Per Capita Income

```{r, warning=F, message=F }
col.name <- c("State", "County", "Tract", "Transit", "Unemployed", "Income")
colnames(census.dat) <- col.name
```

##Format census data

Convert census tract to match format in the houses dataset.

Census tract returned in the api is a six character string ("000501").

```{r, warning=F, message=F, echo=F }
head(census.dat$Tract)
```

But the fips codes matched in the spatial join is an eleven character string ("36067000501").

```{r, warning=F, message=F, echo=F }
head(dat$fips)
```

Append "36067" to the census data tract fields

```{r, warning=F, message=F }
census.dat$Tract <- paste("36067", census.dat$Tract, sep="")
```

New census tract codes:

```{r, warning=F, message=F, echo=F }
head(census.dat$Tract)
```

##Join datsets

Match census data to homes dataset using FIPS codes and store matching census data to houses.

```{r, warning=F, message=F }
tract.order <- match(dat$fips, census.dat$Tract)
dat$transit <- census.dat$Transit[tract.order]
dat$unemployed <- census.dat$Unemployed[tract.order]
dat$income <- census.dat$Income[tract.order]
```

#Aggregate Crime

Load crime data from csv file.

```{r, warning=F, message=F }
crime.dat <- read.csv( "Data/Crime_Locations.csv" )
head(crime.dat)
```

Convert to spatial points and perform spatial join to census tract shapefiles.

```{r, warning=F, message=F }
crime.coord <- crime.dat[ ,c("lon","lat")]
crime <- SpatialPoints(crime.coord, proj4string=CRS("+proj=longlat +datum=WGS84"))
tracts.crime <- over(crime, syr)  
```

Aggregate crimes by census tract.

```{r, warning=F, message=F }
tracts.crime$count <- rep.int(1, 757)
crime.agg <- aggregate(tracts.crime$count, list(tracts.crime$GEOID), FUN="sum")
names(crime.agg) <- c("FIPS", "Crimes")
head(crime.agg)
```

Join aggegrated crime data to homes datset

```{r, warning=F, message=F }
tract.order <- match(dat$fips, crime.agg$FIPS)
dat$crimes <- crime.agg$Crimes[tract.order]
```

Prepare plot colors by crime.

```{r, warning=F, message=F }
color.range <- colorRampPalette( c("steelblue4","light gray","firebrick4") )
color.ramp <- color.range( 10 )
crime.color <- cut(rank(crime.agg$Crimes), breaks=10, labels=color.ramp)
crime.color <- as.character(crime.color)
crime.order <- match(syr$GEOID, crime.agg$FIPS)
crime.color <- crime.color[crime.order]
```

Plot crimes by census tract and house locations.

```{r, warning=F, message=F, fig.width = 10 }
plot(syr, border="white", col=crime.color, main="Crime in the Syracuse Area", xlim=c(-76.2, -76.1), ylim=c(43.0, 43.1))
points(dat$coords, pch=20, cex = 0.9, col="darkorange1")

legend.ranges <- c("1 - 5","6 - 10","11 - 14","15 - 19","20 - 23","24 - 27","28 - 32","33 - 36","37 - 41","42 - 45")
legend("bottomright", bg="white", pch=19, pt.cex=1.5, cex=0.7, legend=legend.ranges, col=color.ramp)
```