---
title: "Data Driven Management II - Lab 6"
author: "Kyle Crichton"
date: "March 8, 2017"
output: github_document
---

#Setup

Load Packages

```{r, warning=F, message=F }
library(maptools)
library(sp)
library(rgeos)
library(raster)
library(RCurl)
library(ggmap)
library(dplyr)
library(rgdal)
```

#Part I

##Download Tiger Shapefiles

Dowloads zip of census track shapefiles, unzips and removes zipped file.

```{r, warning=F, message=F }
download.file("ftp://ftp2.census.gov/geo/tiger/TIGER2010/TRACT/2010/tl_2010_36067_tract10.zip", "Support Files/onondaga census tracts.zip")

unzip("Support Files/onondaga census tracts.zip", exdir = "Support Files/CensusTracks")

file.remove("Support Files/onondaga census tracts.zip")
```

Dowloads zip of primary and secondary road shapefiles, unzips and removes zipped file.

```{r, warning=F, message=F }
download.file("ftp://ftp2.census.gov/geo/tiger/TIGER2015/PRISECROADS/tl_2015_36_prisecroads.zip", "Support Files/primary_and_secondary_roads.zip" )

unzip("Support Files/primary_and_secondary_roads.zip", exdir = "Support Files/Roads")

file.remove("Support Files/primary_and_secondary_roads.zip")
```

##Load and Plot Shapefiles

Loads shapefiles for census tracks and roads.

```{r, warning=F, message=F }
syr <- readShapePoly(fn="Support Files/CensusTracks/tl_2010_36067_tract10", proj4string=CRS("+proj=longlat +datum=WGS84"))

roads <- readShapeLines(fn="Support Files/Roads/tl_2015_36_prisecroads", proj4string=CRS("+proj=longlat +datum=WGS84"))
```

Plot census tracks and roads.

```{r, warning=F, message=F }
par(mar=c(0,0,0,0))
plot(syr, col="grey90" , border="grey40")
plot(roads, col="steelblue", lwd=1, add=T)
```

## Extract Interstates

Extracts interstates from clipped roads shapefile.

```{r, warning=F, message=F }
interstate <- roads[roads$RTTYP == "I", ]
```

Highlight interstates on plot.

```{r, warning=F, message=F }
plot(syr, col="grey90" , border="grey40")
plot(roads, col="steel blue", lwd=1, add=T)
plot(interstate, col="red", lwd=1, add=T)
```


## Clip Interstates

Creates a bounding polygon by mering the census track shapefiles.  In other words, creates an outer border which to clip the roads.

```{r, warning=F, message=F }
border <- gUnaryUnion(syr)
plot(border, col="white" , border="grey40")
plot(roads, col="steelblue", lwd=1, add=T)
```

Clips road shapefiles to fit within the census track border.

```{r, warning=F, message=F }
roads.clipped <- gIntersection(interstate, border)
```

Plot census tracks and roads again to show clipping.

```{r, warning=F, message=F }
plot(syr, col="grey90" , border="grey40")
plot(roads.clipped, col="red", lwd=1, add=T)
```

## Create Buffer

Generates a buffer of approximately a quarter mile around each of the interstates.

```{r, warning=F, message=F }
buffer <- gBuffer(roads.clipped, width = .004)
```

Plot buffer

```{r, warning=F, message=F }
par(mar=c(0,0,0,0))
plot(syr, col="grey90" , border="grey40")
plot(roads.clipped, col="red", lwd=1, add=T)
plot(buffer, border="darkorange3", add=T)
```

##Housing Data

Store URL for data, read in as a csv and store data as  data frame.

```{r, warning=F, message=F }
my.url <- "https://docs.google.com/spreadsheets/d/1W0vM5sCRhZjkQh6A0WGV8j1rhQAjecCrQW7PguHft-E/pub?gid=1989082857&single=true&output=csv"
dat.raw <- getURL( my.url, ssl.verifypeer=FALSE )
dat <- read.csv( textConnection(dat.raw), stringsAsFactors=FALSE )
```

Rename data frame columns

```{r, warning=F, message=F }
names(dat) <- c("timestamp","price","X1","X2","sqft","your.name","lot.size","beds","bath","garage","year","elementary","middle","high","walk","tax","highway","restaurant","starbucks","park","mall","address","zip","tract" )
```

###Format Dataset for Analysis

Remove commas from strings and convert to numerics.

```{r, warning=F, message=F}
dat$price <- as.numeric( gsub( ",","", dat$price ) )
dat$tax <- as.numeric( gsub( ",","", dat$tax ) )
dat$lot.size <- as.numeric( gsub( ",","", dat$lot.size ) )
dat$sqft <- as.numeric( gsub( ",","", dat$sqft ) )
dat$lot.size[ is.na( dat$lot.size ) ] <- mean( dat$lot.size, na.rm=T )
```

###Cleanse Address Information

Some of the addresses stored in the address field are just street addresses, but some contain the city, state and zipcode in addition to the street address.  We want complete street addresses when geocoding the data but in order to standardize the formats we will first remove the extra information from the fields before adding the city, state and zipcode back in.

Example of the unstandardized data:

```{r, warning=F, message=F, echo=F }
dat[122:126, "address"]
```

Remove the strings ", Syracuse NY", ", Syracuse, NY", "Syracuse NY" and "Syracuse, NY"

```{r, warning=F, message=F }
dat$address <- sub(", Syracuse NY", "", dat$address)
dat$address <- sub(", Syracuse, NY", "", dat$address)
dat$address <- sub("Syracuse NY", "", dat$address)
dat$address <- sub("Syracuse, NY", "", dat$address)
```

Remove regular expression patterns matching 5 digits at the end of a string (zipcodes).

```{r, warning=F, message=F }
dat$address <- sub("[0-9]{5}$", "", dat$address)
```

Trim leftover trailing whitespace

```{r, warning=F, message=F }
dat$address <- sub("\\s+$", "", dat$address)
```

Fix bad zipcode (change 1323 to 13203)

```{r, warning=F, message=F }
dat$zip <- sub("1323", "13203", dat$zip)
```

Fixed Data:

```{r, warning=F, message=F, echo=F }
dat[122:126, "address"]
```

###Concatenate Address Fields

Pulls together the standardized address field, appends the city (Syracuse) and state (NY), and then concats the zipcode field.

```{r, warning=F, message=F }
dat$full.addr <- paste(dat$address, ",", sep="")
dat$full.addr <- paste(dat$full.addr, "Syracuse, NY", dat$zip, sep=" ")
```

Sample of full address data:

```{r, warning=F, message=F, echo=F }
dat[122:126, "full.addr"]
```

###Geocode Addresses

```{r, warning=F, message=F }
options( warn=-1 ) 

suppressMessages( dat$coords <- geocode(dat$full.addr, messaging=F ))
```

Plot houses and buffer

```{r, warning=F, message=F }
plot(syr, col="grey90" , border="grey40")
plot(roads.clipped, col="red", lwd=1, add=T)
plot(buffer, border="darkorange3", add=T)
points(dat$coords, pch=20, cex = 0.7, col="darkslategrey")
```

###Add Buffer Flag

Flags houses that fall within the buffer zone.

```{r, warning=F, message=F }
houses <- SpatialPoints(dat$coords, proj4string = CRS("+proj=longlat +datum=WGS84"))
houses.subset <- houses[buffer, ]
dat$interstate <- over(houses, buffer)
dat$interstate[is.na(dat$interstate)] <- 0
```

Highlight houses in buffer zone

```{r, warning=F, message=F }
par(mar=c(0,0,0,0))
plot(syr, col="grey90" , border="grey40")
plot(roads.clipped, col="red", lwd=1, add=T)
points(houses, pch=20, cex = 0.7, col="darkslategrey")
points(houses.subset, pch=20, cex = 0.7, col="darkorange1")
```

#Part II

##GeoJSON Shapefiles

Load geojson shapefiles.

```{r, warning=F, message=F }
syr <- readOGR(dsn = "Support Files/Shapefiles/SYR_GEOJSON.geojson")

syr <- spTransform(syr, CRS=CRS("+proj=longlat +datum=WGS84"))
```

##Industrial Buffer

Generates a buffer of approximately a quarter mile around industrial zones.

```{r, warning=F, message=F }
buffer.industrial <- gBuffer(syr[syr$LandUse == "Industrial", ], width = .004)
```

Plot buffer around industrial zones.

```{r, warning=F, message=F }
color.industrial <- as.character(syr$LandUse)
color.industrial[color.industrial=="Industrial"] <- "yellow4"
color.industrial[color.industrial!="yellow4"] <- "grey90"
par(mar=c(0,0,0,0))
plot(syr, col=color.industrial, border=NA)
plot(buffer.industrial, border="yellow2", add=T)
```

##Add Industrial Flag

Adds a boolean flag to the dataset to indicate proximity to an industrial zone

```{r, warning=F, message=F }
dat$industrial <- over(houses, buffer.industrial)
dat$industrial[is.na(dat$industrial)] <- 0
```

Plot houses inside and outside the industrial buffer zone.

```{r, warning=F, message=F }
color.house.ind <- as.character(dat$industrial)
color.house.ind[color.house.ind=="1"] <- "darkorange1"
color.house.ind[color.house.ind!="darkorange1"] <- "darkslategrey"
par(mar=c(0,0,0,0))
plot(syr, col=color.industrial, border=NA)
plot(buffer.industrial, border="yellow2", add=T)
points(dat$coords, pch=20, cex = 0.7, col=color.house.ind)
```

##School Buffer

Generates a buffer of approximately an eighth of a mile from school.

```{r, warning=F, message=F }
buffer.school <- gBuffer(syr[syr$LandUse == "Schools", ], width = .002)
```

Plot buffer around schools.

```{r, warning=F, message=F }
color.school <- as.character(syr$LandUse)
color.school[color.school=="Schools"] <- "forestgreen"
color.school[color.school!="forestgreen"] <- "grey90"
par(mar=c(0,0,0,0))
plot(syr, col=color.school, border=NA)
plot(buffer.school, border="green3", add=T)
```

##Add School Flag

Adds a boolean flag to the dataset to indicate proximity to a school.

```{r, warning=F, message=F }
dat$school <- over(houses, buffer.school)
dat$school[is.na(dat$school)] <- 0
```

Plot houses inside and outside the school buffer zone.

```{r, warning=F, message=F }
color.house.sch <- as.character(dat$school)
color.house.sch[color.house.sch=="1"] <- "darkorange1"
color.house.sch[color.house.sch!="darkorange1"] <- "darkslategrey"
par(mar=c(0,0,0,0))
plot(syr, col=color.school, border=NA)
plot(buffer.school, border="green3", add=T)
points(dat$coords, pch=20, cex = 0.7, col=color.house.sch)
```